What is an Ancestral Recombination Graph?

### 2022-01-30 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>
Update abstract

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time. This has created exciting
possibilities, with many downstream applications based on
these inferred genealogies. The inferred genetic ancestries are usually
referred to as Ancestral Recombination Graphs, or ARGs. Although initially well
defined as a graphical representation of the coalescent with recombination
stochastic process, the interpretation has become unclear as it is now
also understood to represent a particular realisation of a genetic
ancestry.
Inference methods do not all infer the same information:
although all output marginal trees along the genome, there is a great deal of
variation in how much information of the relationships between the trees is inferred
and retained. An important programme of work over the coming years is to
develop and refine ancestry inference methods, assessing the relative strengths
and weaknesses of the various approaches. Using the blanket term ARG hides the
important differences in the approaches, and hampers our ability to compare and
improve inference methods. In this paper we discuss the different approaches we
may use to encode genetic ancestry, and of the differing amounts of information
about the ancestral process that we can represent and hope to observe. We
classify existing inference methods according to the type of ancestry they
infer, and discuss the strengths and weaknesses of the inferred structure for
downstream applications.
\end{abstract}

@@

In a recombining organism, the genetic ancestry of a longer segment of DNA is 
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing 
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.

There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with 
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.

>>>>

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a representation of the coalescent with recombination
stochastic process as a graph, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individuals monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
\end{abstract}

@@

In a recombining organism, the genetic ancestry of a longer segment of DNA is
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.
%note-jk: could nitpick this in that diamonds don't generate tree "changes" as such. Maybe tweak this on the next round of revision?

% jk-note: this isn't quite right I think, as it's muddling the distinction between a graph stochastic process
% definition with the graph as the encoding of the marginal trees.
There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.
